---
title: "TUD FOR WEEKDAYS"
author: "Kakau with minor amendments from Ben :-)"
date: 'Last run at: `r Sys.time()`'
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: TRUE
---

```{r setup}
# packages ----
library(plyr)
library(stringr)
library(tidyverse)
library(tidyr)
library(lubridate)
library(kableExtra)
library(dplyr)
library(scales)
library(ggplot2)

# parameters ----
dPath <- "/Volumes/hum-csafe/Research Projects/FoliakiPhD/tudData/" # edit for your set up
dataF <- path.expand(paste0(dPath, "TUD Weekdays.csv"))
  
```

#Time Use Diary


Importing Weekends dataframe

The dataframe is a wide format. I had to clean the data by correcting typo and translating some of the data into english. So the file is not as original as it is when export from qualtrics. 

```{r loadData}

# read in my dataframe
weekdays <- read.csv(dataF, na.strings=c(""), check.names = FALSE, header = TRUE, as.is=TRUE)

head(weekdays)



```


Converting the wide dataframe to long dataframe for easier analysis

Ive noticed that the time is in a character format rather than date. i need to change it to date

> see below

```{r makeLong}

longweekdays <- weekdays %>%
  gather(Time, Activity, 4:51)

head(longweekdays)

```


Sorting the TUD dataframe, Firstly, as you can see previously some respondents do not have a secondary activity (i.e NA). Secondly, the data is sorted by time. I think it would be easier to sorted by Household ID. So here i am sorting the data based on ID number

```{r sortByID}
# sorting my dataframe in accending order based on the Household ID
weekdayOrder <- order(longweekdays$`Household ID`)
head(weekdayOrder)

# Applying the sorting order into my "longweekdends" dataframe and call it sortweekends
sortweekdays <- longweekdays[weekdayOrder,]
head(sortweekdays)

names(sortweekdays)
```


Here i am seprating the activity columm into a primary activity and a secondary activity

The issue im facing with is the date. It in character and alsoit picked up the current date (2020) rather than the date from the TUD.

> BA: I can't see a date variable. How would R know which date the diary was completed on?

Im trying to sort the date but it comes back N/A

> see below

```{r rearrangeActs}
# Since both the first and second activiies are in one column, i need to separate them into two columns

#call the number of row in my dataframe for the maximum row for my sequence function
nrow(sortweekdays)

#create a new dataframe "newdata" for the primary activity only (all odd rows)
newdata <- sortweekdays[seq(1,8351,2),]
head(newdata)

#create a new datagame "newdata1" for the secondary activity (oall even rows)
newdata1 <- sortweekdays[seq(2,8352,2),]
head(newdata1)

#since i have separated my two acitvities, i need to merge them into one dataframe
newdata$Activity2 <- newdata1$Activity

kableExtra::kable(head(newdata), caption = "Test merged long form data") %>%
  kable_styling()


```

> BA: fix the time variable so it looks like and acts like HH:MM

```{r fixTime}
newdata$rawTime <- hms::parse_hm(newdata$Time) # this will only be right for AM
newdata$AM_PM <- stringr::word(newdata$Time, 2) # get the second word of 'Time' which will be AM or PM

# Now use that to add 12 hours to rawTime whenit is actually PM
newdata$fixedTime <- ifelse(newdata$AM_PM =="PM", # if PM
                            newdata$rawTime + (12 * 60 * 60), #  add 12 hours
                            newdata$rawTime # otherwise don't
                            ) 
newdata$fixedHMS <- hms::as_hms(newdata$fixedTime) # make sure it is in H:M:S format

t <- table(newdata$Time, newdata$AM_PM)
kableExtra::kable(t, caption = "Test time coding") %>%
  kable_styling()
```

# Weekdays TUD

Here i am trying to seprate the data into days so i can analyse the activities.

```{r createWeekdayDataFrames}

#filter newdata dataframe by date using the dplyr

#creating a new dataframe "Monday"
monday <- dplyr::filter(newdata, Day == "Monday")
head(monday)

#creating a new dataframe "Tuesday"
tuesday <- dplyr::filter(newdata, Day == "Tuesday")
head(tuesday)

#creating a new dataframe "Wednesday"
wednesday <- dplyr::filter(newdata, Day == "Wednesday")
head(wednesday)

#creating a new dataframe "Thursday"
thursday <- dplyr::filter(newdata, Day == "Thursday")
head(thursday)

#creating a new dataframe "Friday"
friday <- dplyr::filter(newdata, Day == "Friday")
head(friday)


```


# Monday

So im trying to sort out the coding for Monday so i can copy it for the rest of the weekdays. I've did a frequency and proportion of activities for Monday.

```{r testMonday}



#frequency of Activities
freqA <- table(monday$Activity, exclude=NULL)
freqA
dim(freqA)

#proportion of Activities
propA <- prop.table(freqA)*100
propA <- round(propA, digits=2)
propA

dim(propA)

#creating table with binding
tableMon <- cbind(freqA, propA)
tableMon <- cbind(Gender = rownames(tableMon), tableMon)
rownames(tableMon) <- NULL
tableMon

#rename column names
colnames(tableMon)[colnames(tableMon) == "freqA"] <- "Frequency"
colnames(tableMon)[colnames(tableMon) == "propA"] <- "Proportion (%)"
tableMon

#using kable extra to create better table format
tableMonday <- kable(tableMon)%>%
  kable_styling(bootstrap_options = "striped", full_width = F)

tableMonday

```


## Monday TUD
Im trying to see the count for all the activities in the TUD

```{r mondayGroupActs}
groupbymonAct <- monday %>%
  dplyr::group_by(Activity) %>%
  dplyr::summarize(count = n())
groupbymonAct

```

Im am trying to remove sleeping, rest, going to work from the data so that i can only focus on the rest of the acitivities.

The issue i am facing as the time is not sorted in accending order. Can you help with the coding

> yep - see what I did :-)

```{r mondayFilter}
tableAll <- monday %>%
  #select(Time, Activity) %>% 
  select(fixedHMS, Activity) %>% # use the new time variable
  filter(Activity != "Sleeping") %>%
  filter(Activity != "Rest") %>%
  filter(Activity != "Work") %>%
  #group_by(Time) %>% 
  group_by(fixedHMS) %>% 
  tally

kableExtra::kable(tableAll, caption = "Monday test (no sleep/rest/work)") %>%
  kable_styling()

# plot
ggplot2::ggplot(tableAll, aes(x = fixedHMS, y = n)) +
  geom_col()
```

## Monday Cooking

Here im am trying to look at anything related to preparing and eating a meal. 

```{r mondayCook}
# creating dataframe for all cooking

tableCooking <- monday %>%
  select(fixedHMS, Activity) %>% 
  filter(Activity != "Sleeping") %>%
  filter(Activity != "Rest") %>%
  filter(Activity != "Work") %>%
  filter(Activity %in% c("Cooking", "Breakfast", "Dinner", "Light meal")) %>% 
  group_by(fixedHMS) %>% 
  tally
  
  
tableCooking

b <- ggplot(tableCooking, aes(x = fixedHMS, y = n)) + geom_bar(stat = "identity", position = "stack", color="blue", fill="blue") +
  ggtitle("(b) Cooking") +
  xlab("Time ") +
  ylab("n") + theme(plot.title = element_text(color="Black", size=11, face="bold.italic", hjust=0.5,lineheight=0.8),
        axis.title.x = element_text(color="black", size=10, face="bold"),
        axis.title.y = element_text(color="black", size=10, face="bold"),
        panel.background=element_rect(fill = "white"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(), 
        axis.line = element_line(colour = "black", size = 1) + 
        scale_x_date(labels = date_format("%I:%M %p"))
)

b

```

## Monday Tidy up

similar to the above but for tidy up/cleaning up
```{r mondayTidyUp}
# creating dataframe for all cleaning up

tableTidyup <- monday %>%
  select(fixedHMS, Activity) %>% 
  filter(Activity != "Sleeping") %>% 
  filter(Activity != "Rest") %>%
  filter(Activity != "Work") %>%
  filter(Activity %in% c("Tidy up", "Laundry", "Showering", "Getting dressed", "Ironing", "Feeding animals", "Turning off appliances/Lights")) %>% 
  group_by(fixedHMS) %>% 
  tally
  
  
tableTidyup

c <- ggplot(tableTidyup, aes(x = fixedHMS, y = n)) + geom_bar(stat = "identity", position = "stack", color="red", fill="red") +
  ggtitle("(c) Tidy up") +
  xlab("Time ") +
  ylab("n") + theme(plot.title = element_text(color="Black", size=11, face="bold.italic", hjust=0.5,lineheight=0.8),
        axis.title.x = element_text(color="black", size=10, face="bold"),
        axis.title.y = element_text(color="black", size=10, face="bold"),
        panel.background=element_rect(fill = "white"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(), 
        axis.line = element_line(colour = "black", size = 1) + 
        scale_x_date(labels = date_format("%I:%M %p"))
)

c

```


## Monday Leisure
  similar to the above but for leisure
```{r momdayLeisure}
# creating dataframe for all leisure

tableleisure <- monday %>%
  select(fixedHMS, Activity) %>% 
  filter(Activity != "Sleeping") %>% 
  filter(Activity != "Rest") %>% 
  filter(Activity != "Work") %>%
  filter(Activity %in% c("TV", "Homework", "Laptop/PC", "Radio", "Mobile phones", "Reading", "Family time")) %>% 
  group_by(fixedHMS) %>% 
  tally
  
  
tableleisure

d <- ggplot(tableleisure, aes(x = fixedHMS, y = n)) + geom_bar(stat = "identity", position = "stack", color="violet", fill="violet") +
  ggtitle("(d) Leisure") +
  xlab("Time ") +
  ylab("n") + theme(plot.title = element_text(color="Black", size=11, face="bold.italic", hjust=0.5,lineheight=0.8),
        axis.title.x = element_text(color="black", size=10, face="bold"),
        axis.title.y = element_text(color="black", size=10, face="bold"),
        panel.background=element_rect(fill = "white"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(), 
        axis.line = element_line(colour = "black", size = 1) + 
        scale_x_date(labels = date_format("%I:%M %p"))
)

d
```

## Monday Away
similar to the above but for away from home
```{r mondayAway}
# creating dataframe for Away from Home

tableAway <- monday %>%
  select(fixedHMS, Activity) %>% 
  filter(Activity != "Sleeping") %>% 
  filter(Activity != "Rest") %>% 
  filter(Activity %in% c("Farm", "Relatives", "Work", "Fitness", "Kava", "Church", "Sports", "Going out"))%>% 
  group_by(fixedHMS) %>% 
  tally
  
  
tableAway

e <- ggplot(tableAway, aes(x = fixedHMS, y = n)) + geom_bar(stat = "identity", position = "stack", color="orange", fill="orange") +
  ggtitle("(e) Away") +
  xlab("Time ") +
  ylab("n") + theme(plot.title = element_text(color="Black", size=11, face="bold.italic", hjust=0.5,lineheight=0.8),
        axis.title.x = element_text(color="black", size=10, face="bold"),
        axis.title.y = element_text(color="black", size=10, face="bold"),
        panel.background=element_rect(fill = "white"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(), 
        axis.line = element_line(colour = "black", size = 1) + 
        scale_x_date(labels = date_format("%I:%M %p"))
)

e
```

## Monday Transport
similar to the above but for transportation

> BA: table breaks for some reason?

```{r mondayTransport}
# creating dataframe for transportation

tableTransport <- monday %>%
  select(fixedHMS, Activity) %>% 
  filter(Activity != "Sleeping") %>%
  filter(Activity != "Rest") %>%
  filter(Activity != "Work") %>%
  filter(Activity %in% c("Public transport", "Driving", "Walking")) %>% 
  group_by(fixedHMS) %>% 
  tally
  
  
tableTransport

#ploting Bar graph

f <- ggplot(tableTransport, aes(x = fixedHMS, y = n)) + geom_bar(stat = "identity", position = "stack", color="purple", fill="purple") +
  ggtitle("(e) Travelling") +
  xlab("Time ") +
  ylab("n") + theme(plot.title = element_text(color="Black", size=11, face="bold.italic", hjust=0.5,lineheight=0.8),
        axis.title.x = element_text(color="black", size=10, face="bold"),
        axis.title.y = element_text(color="black", size=10, face="bold"),
        panel.background=element_rect(fill = "white"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(), 
        axis.line = element_line(colour = "black", size = 1) + 
        scale_x_date(labels = date_format("%I:%M %p"))
)

f

```

## Monday Multiple plots

I wanted to put all the graphs in one page

```{r gridTheGraphs}
#graphs in one page

library(grid)
library(gridExtra)
library(gridBase)


```

> BA: I would do this a different way by re-coding and putting all the plots in one plot:

```{r onePlot}
# I would use data.table for this, not because it is the best way but because I know how to use it!
library(data.table)
mondayDT <- data.table::as.data.table(monday)
mondayDT[, ba_act := ifelse(Activity %in% c("Cooking", "Breakfast", "Dinner", "Light meal"), 
                            "Cooking", # <- new code
                            NA)
         ]

mondayDT[, ba_act := ifelse(Activity %in% c("Tidy up", "Laundry", "Showering", "Getting dressed", "Ironing", "Feeding animals", "Turning off appliances/Lights"),
                            "Tidy up", 
                            ba_act)
         ]

mondayDT[, ba_act := ifelse(Activity %in% c("TV", "Homework", "Laptop/PC", "Radio", "Mobile phones", "Reading", "Family time"), 
                            "Leisure", 
                            ba_act)
         ]

mondayDT[, ba_act := ifelse(Activity %in% c("Farm", "Relatives", "Work", "Fitness", "Kava", "Church", "Sports", "Going out"), 
                            "Away", 
                            ba_act)
         ]

mondayDT[, ba_act := ifelse(Activity %in% c("Public transport", "Driving", "Walking"), 
                            "Transport", 
                            ba_act)
         ]

t <- table(mondayDT$Activity, mondayDT$ba_act, useNA = "always")

kableExtra::kable(t, caption = "Test activity coding") %>%
  kable_styling()

# make a summary table for the plot
plotDT <- mondayDT[!is.na(ba_act), # excludes all NAs - see table to check coding 
                   .(count = .N), keyby = .(fixedHMS, ba_act)]

ggplot2::ggplot(plotDT, aes(x = fixedHMS, y = count, fill = ba_act)) +
  geom_col() +
  scale_fill_discrete(name="Activity (filtered)")
```

# All weekdays plotted using facets

Figure \@ref(fig:allWeekdays) shows counts for all days. We can see a few days are unkown. Also we have higher counts on Mondays. A lot of people are 'out' late at night?

```{r allWeekdays, fig.cap="All days"}
# use the all days table and the same coding method but across all days

allDaysDT <- data.table::as.data.table(newdata)
allDaysDT[, ba_act := ifelse(Activity %in% c("Cooking", "Breakfast", "Dinner", "Light meal"), 
                            "Cooking", # <- new code
                            NA)
         ]

allDaysDT[, ba_act := ifelse(Activity %in% c("Tidy up", "Laundry", "Showering", "Getting dressed", "Ironing", "Feeding animals", "Turning off appliances/Lights"),
                            "Tidy up", 
                            ba_act)
         ]

allDaysDT[, ba_act := ifelse(Activity %in% c("TV", "Homework", "Laptop/PC", "Radio", "Mobile phones", "Reading", "Family time"), 
                            "Leisure", 
                            ba_act)
         ]

allDaysDT[, ba_act := ifelse(Activity %in% c("Farm", "Relatives", "Work", "Fitness", "Kava", "Church", "Sports", "Going out"), 
                            "Away", 
                            ba_act)
         ]

allDaysDT[, ba_act := ifelse(Activity %in% c("Public transport", "Driving", "Walking"), 
                            "Transport", 
                            ba_act)
         ]

# t <- table(allDaysDT$Activity, allDaysDT$ba_act, useNA = "always")
# 
# kableExtra::kable(t, caption = "Test activity coding (all days)") %>%
#   kable_styling()

# make a summary table for the plot
plotDT <- allDaysDT[!is.na(ba_act), # excludes all NAs - see table to check coding 
                   .(count = .N), keyby = .(fixedHMS, ba_act, Day)]

ggplot2::ggplot(plotDT, aes(x = fixedHMS, y = count, fill = ba_act)) +
  geom_col() +
  facet_grid(Day ~ .) +
  scale_fill_discrete(name="Activity (filtered)")
```